The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-06 kalman_walk_paper.tex
edited apx

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A, B, C) 
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &:= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  .
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}

@@

We could use this to compute the gradient of $D$,
or to get the quadratic approximation to $D$ near the optimal set.
To do so, it'd be nice to have a way to compute the inner integral above.
Suppose that we can diagonalize $A = U \Lambda U^{-1}$.
Then
\begin{equation} \label{eqn:exp_deriv}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
  \end{aligned}
\end{equation}
Now, notice that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
      &=
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
  \end{aligned}
\end{equation}
Therefore, 
defining
\begin{equation}
    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
\end{equation}
moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      U X(t,Z) U^{-1} .
  \end{aligned}
\end{equation}

Following on from above, we see that if $Z=\Delta_{k \ell}$, then
\begin{equation}
  \begin{aligned}
      X_{ij}^{k\ell}(t) = 
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
      (U^{-1})_{\cdot k} U_{\ell \cdot},
  \end{aligned}
\end{equation}
where $U_{k \cdot}$ is the $k$th row of $U$,
and so
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &=
        \int_0^\infty
            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
        dt .
    \end{aligned}
\end{equation}
This implies that
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)
        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
    \end{aligned}
\end{equation}
and so
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)
        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
    \end{aligned}
\end{equation}

@@

More generally, $B$ and $C$ may also change.
To extend this we need the remaining second derivatives of $D$.
First, in $B$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)
        (uY+vZ)^T
        \vert_{u=v=0} 
        e^{t A_0^T} C_0^T 
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0^T} C_0^T 
        \right\} dt  .
  \end{aligned}
\end{equation}
Next, in $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)^T
        (uY+vZ)
        \vert_{u=v=0} 
        e^{t A_0} B_0
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
Now, the mixed derivatives in $B$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
In $A$ and $B$
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        Z^T e^{t A_0} C_0
        \right\} dt  ,
  \end{aligned}
\end{equation}
and finally in $A$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        B_0 e^{t A_0} Z
        \right\} dt  .
  \end{aligned}
\end{equation}

@@

%%%%%%%%%%%
\section*{Evolution of segregation covariance}
\label{apx:seg_cov}

>>>>

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A, B, C)^2
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &:= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  ,
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
where $\tr X$ denotes the trace of a square matrix $X$.
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)^2 \vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so without loss of generality, $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)^2 \vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}

@@

We could use this to get the quadratic approximation to $D$ near the optimal set.
To do so, it'd be nice to have a way to compute the inner integral above.
Suppose that we diagonalize $A = U \Lambda U^{-1}$.
Then
\begin{equation} \label{eqn:exp_deriv}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      \int_0^t U e^{\Lambda s} U^{-1} Z U e^{\Lambda (t-s)} U^{-1} ds 
  \end{aligned}
\end{equation}
Now, notice that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{s \lambda_i} e^{(t-s) \lambda_j} ds
      &=
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } .
  \end{aligned}
\end{equation}
Therefore, 
defining
\begin{equation}
    X_{ij}(t,Z) = \left( U^{-1} Z U \right)_{ij}
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
\end{equation}
moving the $U$ and $U^{-1}$ outside the integral and integrating we get that
\begin{equation}
  \begin{aligned}
      \int_0^t e^{As} Z e^{A(t-s)} ds 
      &=
      U X(t,Z) U^{-1} .
  \end{aligned}
\end{equation}

Following on from above, we see that if $Z=\Delta_{k \ell}$, then
\begin{equation}
  \begin{aligned}
      X_{ij}^{k\ell}(t) = 
      \frac{ e^{t \lambda_i} - e^{t \lambda_j} }{ \lambda_i - \lambda_j } 
      (U^{-1})_{\cdot k} U_{\ell \cdot},
  \end{aligned}
\end{equation}
where $U_{k \cdot}$ is the $k$th row of $U$,
and so
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &=
        \int_0^\infty
            \rho(t) \tr\left\{ C U X^{ij}(t) U^{-1} B B^T (U^{-1})^T X^{k\ell}(t)^T U^T C^T \right\}
        dt .
    \end{aligned}
\end{equation}
This implies that
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)^2
        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
    \end{aligned}
\end{equation}
and so
\begin{equation}
    \begin{aligned}
        D(A_0+\epsilon Z)^2
        &\approx \epsilon^2\sum_{ijk\ell} H^{ij,k\ell} Z_{ij} Z_{k\ell} 
    \end{aligned}
\end{equation}

@@

More generally, $B$ and $C$ may also change.
To extend this we need the remaining second derivatives of $D^2$.
First, in $B$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY+vZ,C_0)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)
        (uY+vZ)^T
        \vert_{u=v=0} 
        e^{t A_0^T} C_0^T 
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        C_0 e^{t A_0}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0^T} C_0^T 
        \right\} dt  .
  \end{aligned}
\end{equation}
Next, in $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0,C_0+uY+vZ)\vert_{u=v=0} \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \frac{d}{du} \frac{d}{dv} 
        (uY+vZ)^T
        (uY+vZ)
        \vert_{u=v=0} 
        e^{t A_0} B_0
        \right\} dt  \\
        &=
        \frac{1}{2} \int_0^\infty \rho(t) \tr\left\{ 
        B_0 e^{t A_0^T}
        \left( Y Z^T + Z Y^T \right)
        e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
Now, the mixed derivatives in $B$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{B,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0,B_0+uY,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        Y e^{t A_0^T} C_0^T Z e^{t A_0} B_0
        \right\} dt  .
  \end{aligned}
\end{equation}
In $A$ and $B$
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,B}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0+vZ,C_0)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        Z^T e^{t A_0} C_0
        \right\} dt  ,
  \end{aligned}
\end{equation}
and finally in $A$ and $C$:
\begin{equation}
    \begin{aligned}
        \mathcal{H}^{A,C}(Y,Z)
        &:= 
        \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY,B_0,C_0+vZ)\vert_{u=v=0} \\
        &=
        \int_0^\infty \rho(t) \tr\left\{ 
        C_0 \left(\int_0^t e^{s A_0} Y e^{(t-s) A_0} ds \right) B_0
        B_0 e^{t A_0} Z
        \right\} dt  .
  \end{aligned}
\end{equation}

Together, numerical computation of these expressions,
along with estimates of genetic covariance within a population,
allow precise predictions of evolutionary dynamics of a particular system.
The approximation should be good as long as the second-order Taylor approximation holds.

@@

%%%%%%%%%%%
\section{Evolution of segregation covariance}
\label{apx:seg_cov}
